---
title: "United Airlines Departure Delays"
authors: Dan Nguyen, Jennifer Poling, Paul Skentzos
output: html_document
date: "2025-10-25"
---
# United Airlines Departure Delays Project

This project will examine the relationship between United Airlines departure delays and time of day, time of year, temperature, wind speed, precipitation, and visibility.

## Introduction
We set out to understand when and under what conditions United flights tend to depart late from New York–area airports, using the public nycflights13 dataset for 2013, imported in our RStudio environment. We look at time of day, time of year, and common weather factors: temperature, wind speed, precipitation, and visibility. We summarize what the data shows. To check whether visible patterns are likely more than chance, we use permutation tests, a resampling method that shuffles labels and compares the observed result to what we’d expect if there were no real relationship. 

###### Install libraries
```{r}
library(tidyverse)
library(nycflights13)
```
## Prepare the Data

###### Inspect the data
```{r}
glimpse(flights)
```
```{r}
glimpse(weather)
```

###### Keep only the relevant United Airlines flights 

```{r}
flights_UA <- flights %>%
  filter(carrier == 'UA') %>%
  filter(!is.na(dep_delay)) 
```


###### Create the "late" and "very_late" columns
```{r}
flights_UA <- flights_UA %>%
  mutate(
    late = dep_delay > 0,
    very_late = dep_delay > 30
    )
glimpse(flights_UA)
```

###### Join the flight data and the weather data.

Use a left join so that the weather data is applied to each flight

```{r}
flights_weather <- flights_UA %>%
  left_join(weather, by = c('origin', 'time_hour', 'year', 'month', 'day', 'hour'))
glimpse(flights_weather)
```

###### Create a "season" column.
First check how many are null values in the "month" column.
```{r}
sum(is.na(flights_weather$month))
```
Now create the "season" column. Use this definition: summer consists of June, July, and August; fall consists of September, October, and November; winter consists of December, January, and February; spring consists of March, April, and May.
```{r}
flights_weather <- flights_weather %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ 'winter',
    month %in% c(3, 4, 5) ~ 'spring',
    month %in% c(6, 7, 8) ~ 'summer',
    month %in% c(9, 10, 11) ~ 'fall'
  ))
glimpse(flights_weather)
```

###### Create a time_of_day column.
What times do flights operate?
```{r}
unique(flights_weather$hour)
```
How many null values are there in the hour column?
```{r}
sum(is.na(flights_weather$hour))
```
Create a time_of_day column using the following definition. early: 5, 6, 7; morning: 8, 9, 10; mid_day: 11, 12, 13; afternoon: 14, 15, 16; evening: 17, 18, 19; late: 20, 21, 22, 23
```{r}
flights_weather <- flights_weather %>%
  mutate(time_of_day = case_when(
    hour %in% c(5, 6, 7) ~ 'early',
    hour %in% c(8, 9, 10) ~ 'morning',
    hour %in% c(11, 12, 13) ~ 'mid_day',
    hour %in% c(14, 15, 16) ~ 'afternoon',
    hour %in% c(17, 18, 19) ~ 'evening',
    hour %in% c(20, 21, 22, 23) ~ 'night' 
  ))
glimpse(flights_weather)    
```
######## Convert the time_of_day column to a factor with a specific order for consistent plotting and analysis
```{r}
flights_weather <- flights_weather %>%
  mutate(time_of_day = fct_relevel(time_of_day,
                                 'early',
                                 'morning',
                                 'mid_day',
                                 'afternoon',
                                 'evening',
                                 'night'))
count(flights_weather, time_of_day)
```

###### Create Visibility Categories
What is in the visibility column?
```{r}
unique(flights_weather$visib)
```
How many NA values are there in the visibility column?
```{r}
sum(is.na(flights_weather$visib))
```
Keep the rows that have visibility values
```{r}
flights_weather <- flights_weather %>%
  filter(!is.na(visib))
```


Create visibility categories based on common visibility cutoffs in aviation. IMC (Instrument Meteorological Conditions): below 3, call this 'poor'. MVFR (Marginal Visual Flight Rules): 3 to 5, call this 'fair'. VFR (Visual Flight Rules): above 5, call this 'good'.
```{r}
flights_weather <- flights_weather %>%
  mutate(visib_cat = case_when(
    visib < 3 ~ 'poor',
    visib >=3 & visib <= 5 ~ 'fair',
    visib > 5 ~ 'good'
  ))
glimpse(flights_weather)
```

######## Convert the visibility category column to a factor with a specific order for consistent plotting and analysis
```{r}
flights_weather <- flights_weather %>%
  mutate(visib_cat = fct_relevel(visib_cat,
                                 'poor',
                                 'fair',
                                 'good'))
count(flights_weather, visib_cat)
```
###### Create a column for whether or not there is precipitation.
```{r}
flights_weather <- flights_weather %>%
  mutate(precip_binary = precip > 0.00)
count(flights_weather, precip_binary)
```
###### Create quantile bins for temperature.
```{r}
sum(is.na(flights_weather$temp))
```
Remove rows with NA in the temperature column.
```{r}
flights_weather <- flights_weather %>%
  filter(!is.na(temp))
```
Calculate the quantiles
```{r}
temp_quantiles <- quantile(flights_weather$temp, probs = seq(0, 1, 0.25))
```
Create a new column
```{r}
flights_weather$temp_q <- cut(flights_weather$temp,
                              breaks = temp_quantiles,
                              include.lowest = TRUE,
                              right = FALSE)

# Check the distribution of the new bins
table(flights_weather$temp_q)
```
###### Create quantile bins for wind speed.
```{r}
sum(is.na(flights_weather$wind_speed))
```
Remove rows with NA in the wind speed column.
```{r}
flights_weather <- flights_weather %>%
  filter(!is.na(wind_speed))
```
Calculate the quantiles
```{r}
wind_speed_quantiles <- quantile(flights_weather$wind_speed, probs = seq(0, 1, 0.25))
```
Create a new column
```{r}
flights_weather$wind_speed_q <- cut(flights_weather$wind_speed,
                              breaks = wind_speed_quantiles,
                              include.lowest = TRUE,
                              right = FALSE)

# Check the distribution of the new bins
table(flights_weather$wind_speed_q)
```
###### Create a column for whether or not there are wind gusts.
I think that NA means there weren't wind gusts. Check how many NA values in the wind gust column.
```{r}
sum(is.na(flights_weather$wind_gust))
```
Create a wind gust binary column.
```{r}
flights_weather <- flights_weather %>%
  mutate(wind_gust_binary = !is.na(wind_gust))
count(flights_weather, wind_gust_binary)
```
Create a summary table.
```{r}
summary(flights_weather)
```
Output the clean .csv
```{r}
write_csv(flights_weather, 'flights_clean.csv')
```

###### Exploratory Data Analysis  
####### Departure delay vs time of day and time of year  
```{r}
ggplot(flights_weather, aes(x = time_of_day, y = dep_delay)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.9) +
  coord_cartesian(ylim = c(quantile(flights_weather$dep_delay, 0.01, na.rm=TRUE),
                           quantile(flights_weather$dep_delay, 0.99, na.rm=TRUE))) +
  facet_wrap(~season) +
  labs(title = "Departure delay by time of day (faceted by season)",
       x = "Time of day", y = "Departure delay (minutes)") +
  theme_minimal(base_size = 13) + theme(axis.text.x = element_text(angle = 70, vjust = 0.5))
```
######## Q1: Do flights departing in the evening and at night tends to get delayed longer than flights departing at earlier time?  
H~0: $\mu$~n - $\mu$~e = 0  
H~A: $\mu$~n - $\mu$~e $\ge$ 0  
```{r}
evening_night = flights_weather$dep_delay[flights_weather$time_of_day == "evening" | flights_weather$time_of_day == "night"]
earlier = flights_weather$dep_delay[flights_weather$time_of_day != "evening" & flights_weather$time_of_day != "night"]

Time <- c(evening_night, earlier)

# Observed difference in means
observed <- mean(evening_night) - mean(earlier)
observed

# Get the sample sizes
n1 <- length(evening_night)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```
Based on the p-value, we reject the null hypothesis. Flights departing in the evening and at night tends to get delayed longer than flights departing at earlier time.  
######## Q2: Do flights departing in the Holidays Season (Winter & Summer) tends to get delayed longer than flights departing in other seasons?  
H~0: $\mu$~H - $\mu$~O = 0  
H~A: $\mu$~H - $\mu$~O $\ge$ 0  
```{r}
holiday = flights_weather$dep_delay[flights_weather$season == "winter" | flights_weather$season == "summer"]
other = flights_weather$dep_delay[flights_weather$season == "spring" & flights_weather$season != "fall"]

Time <- c(holiday, other)

# Observed difference in means
observed <- mean(holiday) - mean(other)
observed

# Get the sample sizes
n1 <- length(holiday)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```
Based on the p-value, we reject the null hypothesis. Flights departing in the holidays season to get delayed slightly longer than flights departing in other seasons.  

####### Departure delay vs Temperature
```{r}
ggplot(flights_weather, aes(x = temp, y = dep_delay)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(
    title = "Relationship between Temperature and Departure Delay",
    x = "Temperature (°F)",
    y = "Departure Delay (minutes)"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
flights_weather %>%
  group_by(temp_q) %>%
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE),
            median_delay = median(dep_delay, na.rm = TRUE),
            n = n()) %>%
  ggplot(aes(x = temp_q, y = mean_delay)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(mean_delay, 1)), vjust = -0.3, size = 3) +
  labs(
    title = "Average Departure Delay by Temperature (4 quantiles)",
    x = "Temperature bin (°F)",
    y = "Mean dep_delay (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
ggplot(flights_weather, aes(x = factor(temp_q), y = dep_delay)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.7) +
  labs(
    title = "Departure Delay Distribution across Temperature Quartiles",
    x = "Temperature quartile",
    y = "Departure delay (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```
######## Q1: Do flights departing in warmer temperatures (4th quartile) tends to get delayed longer than flights departing in colder temperature (1st quartile)?  
H~0: $\mu$~W - $\mu$~C = 0  
H~A: $\mu$~W - $\mu$~C $\ge$ 0  

```{r}
cold = flights_weather$dep_delay[flights_weather$temp >= 10.9 & flights_weather$temp < 42.1]
warm = flights_weather$dep_delay[flights_weather$temp >= 73 & flights_weather$temp <= 100]

Time <- c(warm, cold)

# Observed difference in means
observed <- mean(warm) - mean(cold)
observed

# Get the sample sizes
n1 <- length(warm)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```
####### Departure delay vs Wind Speed
```{r}
ggplot(flights_weather, aes(x = wind_speed, y = dep_delay)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(
    title = "Relationship between Temperature and Departure Delay",
    x = "Wind Speed",
    y = "Departure Delay (minutes)"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
flights_weather %>%
  group_by(wind_speed_q) %>%
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE),
            median_delay = median(dep_delay, na.rm = TRUE),
            n = n()) %>%
  ggplot(aes(x = wind_speed_q, y = mean_delay)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(mean_delay, 1)), vjust = -0.3, size = 3) +
  labs(
    title = "Average Departure Delay by Wind Speed (4 quantiles)",
    x = "Wind Speed (mph)",
    y = "Mean dep_delay (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
######## Q1: Do flights departing in stronger wind (4th quartile) tends to get delayed longer than flights departing in milder wind (1st quartile)?  
H~0: $\mu$~F - $\mu$~S = 0  
H~A: $\mu$~F - $\mu$~S $\ge$ 0  

```{r}
mild = flights_weather$dep_delay[flights_weather$wind_speed >= 0 & flights_weather$wind_speed < 6.9]
strong = flights_weather$dep_delay[flights_weather$wind_speed >= 13.8 & flights_weather$wind_speed <= 42.6]

Time <- c(strong, mild)

# Observed difference in means
observed <- mean(strong) - mean(mild)
observed

# Get the sample sizes
n1 <- length(mild)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```
####### Departure delay vs Precipitation
```{r}
ggplot(flights_weather, aes(x = precip_binary, y = dep_delay, fill = precip_binary)) +
  geom_boxplot(outlier.size = 0.7, alpha = 0.8) +
  scale_fill_manual(values = c("skyblue3", "gray50"),
                    labels = c("No precipitation", "Precipitation")) +
  labs(
    title = "Departure Delay by Precipitation Condition",
    x = "Precipitation (0 = No, 1 = Yes)",
    y = "Departure delay (minutes)",
    fill = "Precipitation"
  ) +
  coord_cartesian(ylim = quantile(flights$dep_delay, c(0.01, 0.99), na.rm = TRUE)) +
  theme_minimal(base_size = 13)
```
######## Q1: Do flights departing in precipitation tends to get delayed longer than flights departing in dry condition?  
H~0: $\mu$~W - $\mu$~D = 0  
H~A: $\mu$~W - $\mu$~D $\ge$ 0  

```{r}
wet = flights_weather$dep_delay[flights_weather$precip_binary == TRUE]
dry = flights_weather$dep_delay[flights_weather$precip_binary == FALSE]

Time <- c(wet, dry)

# Observed difference in means
observed <- mean(wet) - mean(dry)
observed

# Get the sample sizes
n1 <- length(wet)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```
####### Departure delay vs Visibility
```{r}
ggplot(flights_weather, aes(x = visib, y = dep_delay)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  labs(
    title = "Departure Delay vs Visibility",
    x = "Visibility (miles)",
    y = "Departure Delay (minutes)"
  ) +
  theme_minimal()
```


```{r}
ggplot(flights_weather, aes(x = visib_cat, y = dep_delay, fill = visib_cat)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Departure Delay by Visibility Category",
    x = "Visibility Category",
    y = "Departure Delay (minutes)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
######## Q1: Do flights departing in poor visibilty tends to get delayed longer than flights departing in good visibility?  
H~0: $\mu$~P - $\mu$~G = 0  
H~A: $\mu$~P - $\mu$~G $\ge$ 0  

```{r}
poor = flights_weather$dep_delay[flights_weather$visib_cat == 'poor']
good = flights_weather$dep_delay[flights_weather$visib_cat == 'good']

Time <- c(poor, good)

# Observed difference in means
observed <- mean(poor) - mean(good)
observed

# Get the sample sizes
n1 <- length(poor)
n_total <- length(Time)

# Permutation Test
N <- 10^4 - 1 # number of times to repeat this process
result <- numeric(N) # space to save the random differences

for (i in 1:N) {
  index <- sample(n_total, size = n1, replace = FALSE)
  result[i] <- mean(Time[index]) - mean(Time[-index])
}

# Plot histogram
ggplot() + geom_histogram(aes(result), bins = 20) +
geom_vline(xintercept = observed, linetype = "dashed")

# P-value
p_value = (sum(result >= observed) + 1)/(N + 1)
p_value
```

```{r fig-late-by-precip, message=FALSE, warning=FALSE, fig.cap="Proportion of Late Flights by Precipitation Condition"}
# Calculate the proportion of late flights under precipitation vs no precipitation
late_by_precip <- flights_weather %>%
  group_by(precip_binary) %>%
  summarise(late_rate = mean(late, na.rm = TRUE), .groups = "drop")

# Bar plot showing the share of late departures by precipitation
ggplot(late_by_precip, aes(x = precip_binary, y = late_rate, fill = precip_binary)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Late Departures by Precipitation",
    x = "Precipitation (No / Yes)",
    y = "Late Flights (%)"
  ) +
  theme_minimal(base_size = 13)
late_by_precip
```

Each bar shows the percentage of flights that departed late within that weather condition.
The two bars represent conditional probabilities (“chance of a late flight given the condition”), not shares of all flights.
This matches the permutation test that compares the difference in late-flight proportions between the two weather conditions.
---


```{r fig-late-by-visib, message=FALSE, warning=FALSE, fig.cap="Proportion of Late Flights by Visibility Category"}
# Calculate the proportion of late flights by visibility category
late_by_visib <- flights_weather %>%
  group_by(visib_cat) %>%
  summarise(late_rate = mean(late, na.rm = TRUE), .groups = "drop")

# Bar plot showing the share of late departures by visibility
ggplot(late_by_visib, aes(x = visib_cat, y = late_rate, fill = visib_cat)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Late Departures by Visibility",
    x = "Visibility Category",
    y = "Late Flights (%)"
  ) +
  theme_minimal(base_size = 13)
late_by_visib
```
The bars represent the fraction of flights that left late within each visibility category.
Each bar answers a separate question:
Out of all flights that day with this visibility, what share were late?

The result shows that low-visibility periods have a higher late-departure rate, consistent with the permutation test for differences in proportions.

